Linear stability of magnetized massive protoplanetary disks

### 2014-02-21 paper/results.tex
marginal notes

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_ideal}
  \caption{Maximum gravitational potential energy perturbation
    associated with the MRI in a self-gravitating disk with $Q=0.2$ in
    the limit of ideal MHD ($\Lambda_0=10^2$). 
    \label{gravity_energy}}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_ideal_sg}
  \caption{Example of a linear MRI mode with non-zero density and
    potential perturbation, in a strongly magnetized massive disk
    ($\beta=10$ and $Q=0.2$) in the limit of ideal MHD
    ($\Lambda_0=10^2$).   
    \label{mri_massive}}
\end{figure}

\subsubsection{Resistive case}
We repeat the above calculation for resistive disks. We fix
$\beta=100$. Fig. \ref{gravity_energy_resis}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{Maximum gravitational potential energy perturbation
    associated with the MRI in a self-gravitating disk with $Q=0.2$,
    for a range of midplane Elsasser numbers.   
    \label{gravity_energy_resis}}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_resis_sg}
  \caption{Example of a linear resistive MRI mode with significant
    gravitational potential perturbation.  
    \label{mri_massive_resis}}
\end{figure}

@@

We imagine an iterative procedure to solve the above equations,
starting from the \emph{Cowling approximation}, defined to be the
solution to Eq. \ref{simp1}---\ref{simp3} with potential perturbation
set to zero. This is the standard MRI and we denote the solution as
$\dvx^{(0)}$, $\dvy^{(0)}$ and $W^{(0)}$.

We now include self-gravity. In general, the MRI will have some
density perturbation so $W^{(0)}\neq0$. The Poisson equation implies
an associated potential perturbation,    
\begin{align} 
  \dphi = -\frac{\Omega^2}{c_s^2 Q k^2} W^{(0)}.
\end{align}
Physically, we expect $k^2\geq0$, so that a positive (negative) density
perturbation causes a negative (positive) potential perturbation. We
then regard $\dphi$ to alter the Cowling solution, so
that $\dvx^{0} \to \dvx^{0} + \dvx^{(1)}$ and similarly for $\dvy$ and
$W$. Then 
\begin{align}
  &   k_x\sigma \dphi = v_A^2k^2\dvx^{(1)} + \imgi\sigma\left[\imgi\sigma
  \dvx^{(1)} - 2\Omega\dvy^{(1)} + \imgi k_x W^{(1)}\right], \label{simp_pert1}\\ 
  &  0= v_A^2k_z^2\left[\dvy^{(1)} + \frac{\imgi S
    }{\sigma}\dvx^{(1)}\right] + \imgi\sigma\left[\imgi\sigma\dvy^{(1)} +
  \frac{\kappa^2}{2\Omega}\dvx^{(1)}\right],\label{simp_pert2}\\
  & k_z^2\dphi  = \left(\frac{\sigma^2}{c_s^2}-k_z^2\right)W^{(1)} +
  \sigma k_x \dvx^{(1)}.\label{simp_pert3} 
\end{align}
Now, if the perturbations to the magnetic field remain
unchanged, i.e. the mode remains close to the standard MRI, then
$\dvx^{(1)} \sim 0$ and $\dvy^{(1)}\sim0$, so Eq. \ref{simp_pert2} is
satisfied. Eq. \ref{simp_pert1} then require $\dphi +
W^{(1)}\sim0$. This is compatible with Eq. \ref{simp_pert3} if 
\begin{align}
  \left|k_z^2\right| \gg \left|\frac{\sigma^2}{c_s^2}\right|. \label{cond}
\end{align}
For the ideal MRI,  we assume $k_z\sim \Omega/v_A$. Then
$|\sigma^2/c_s^2k_z^2|\sim |\sigma^2/\Omega^2\beta|\ll1$ because
$|\sigma|\lesssim \Omega$ and we are considering $\beta\gtrsim
10$. Thus Eq. \ref{cond} is generally satisfied.

>>>>

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_ideal}
  \caption{Maximum gravitational potential energy perturbation
    associated with the MRI in a self-gravitating disk with $Q=0.2$ in
    the limit of ideal MHD ($\Lambda_0=10^2$), for a range of field
    strengths $\beta$.  
    \label{gravity_energy}}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_ideal_sg}
  \caption{Example of a linear MRI mode with non-negligible density and
    potential perturbation, in a strongly magnetized massive disk
    ($\beta=10$ and $Q=0.2$) in the limit of ideal MHD 
    ($\Lambda_0=10^2$).   
    \label{mri_massive}}
\end{figure}

\subsubsection{Resistive case}
We repeat the above calculation for resistive disks, but fix
$\beta=100$. Fig. \ref{gravity_energy_resis} shows the maximum
perturbed gravitational energy increases with 
resistivity. Interestingly, in the highly resistive case
the MRI can be energeically dominated by self-gravity. An example is shown in
Fig. \ref{mri_massive_resis}, which  plots the energies assocaited with the 
linear MRI mode with $k_xH\sim1.3$ in the $\Lambda_0=0.1$  
disk. The gravitational potential energy significantly exceeds the
magnetic energy near the midplane ($z\lesssim0.3H$).

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{Maximum gravitational potential energy perturbation
    associated with the MRI in a self-gravitating disk with $Q=0.2$,
    for a range of midplane Elsasser numbers $\Lambda_0$ at fixed
    field strength $\beta=100$.    
    \label{gravity_energy_resis}}
\end{figure}

We also calculated the mode in Fig. \ref{mri_massive_resis} with the \emph{Cowling 
  approximation}, where the Poisson equation is ignored in the 
linearized equations and the potential perturbation set to zero  
(formally, we set $Q\to\infty$ in Eq. \ref{lin_poisson}).  
We compare the magnetic energy in the Cowling approximation to that of
the full problem in Fig. \ref{mri_massive_cowling}. The growth rate
increases when self-gravity is included in the linear response, as
expected, but only slightly. $\gamma$ and $E_m$ are similar in both cases,
indicating that the instability in the self-gravitating calculation is
still due to the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_resis_sg}
  \caption{Example of a linear resistive MRI mode with significant
    gravitational potential energy perturbation.  The vertical line
    inidicates $\Lambda=1$.  
    \label{mri_massive_resis}}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_result_cowling}
  \caption{Magnetic energy associated with the linear mode shown in
    Fig. \ref{mri_massive_resis} (solid) compared with that computed
    under the Cowling approximation (dotted). 
    \label{mri_massive_cowling}}
\end{figure}

@@

We imagine an iterative procedure to solve the above equations,
starting from the Cowling approximation. %defined to be the
%solution to Eq. \ref{simp1}---\ref{simp3} with potential perturbation
%set to zero. 
This is the standard MRI and we denote the solution as
$\dvx^{(0)}$, $\dvy^{(0)}$ and $W^{(0)}$.

We now include self-gravity. In general, the compressible MRI has
$W^{(0)}\neq0$. The Poisson equation implies
an associated potential perturbation,    
\begin{align} 
  \dphi = -\frac{\Omega^2}{c_s^2 Q k^2} W^{(0)}.
\end{align}
Physically, we expect $k^2\geq0$, so that a positive (negative) density
perturbation causes a negative (positive) potential perturbation. We
then regard $\dphi$ to alter the Cowling solution, so
that $\dvx^{0} \to \dvx^{0} + \dvx^{(1)}$ and similarly for $\dvy$ and
$W$. Then 
\begin{align}
  &   k_x\sigma \dphi = v_A^2k^2\dvx^{(1)} + \imgi\sigma\left[\imgi\sigma
  \dvx^{(1)} - 2\Omega\dvy^{(1)} + \imgi k_x W^{(1)}\right], \label{simp_pert1}\\ 
  &  0= v_A^2k_z^2\left[\dvy^{(1)} + \frac{\imgi S
    }{\sigma}\dvx^{(1)}\right] + \imgi\sigma\left[\imgi\sigma\dvy^{(1)} +
  \frac{\kappa^2}{2\Omega}\dvx^{(1)}\right],\label{simp_pert2}\\
  & k_z^2\dphi  = \left(\frac{\sigma^2}{c_s^2}-k_z^2\right)W^{(1)} +
  \sigma k_x \dvx^{(1)}.\label{simp_pert3} 
\end{align}
Now, if the perturbations to the magnetic field remain
unchanged, i.e. the mode remains close to the standard MRI, then
$\dvx^{(1)} \sim 0$ and $\dvy^{(1)}\sim0$, so Eq. \ref{simp_pert2} is
satisfied. Eq. \ref{simp_pert1} then require $\dphi +
W^{(1)}\sim0$. This is compatible with Eq. \ref{simp_pert3} if 
\begin{align}
  \left|k_z^2\right| \gg \left|\frac{\sigma^2}{c_s^2}\right|. \label{cond}
\end{align}
For the ideal MRI,  we assume $k_z\sim \Omega/v_A$. Then
$|\sigma^2/c_s^2k_z^2|\sim |\sigma^2/\Omega^2\beta|\ll1$ because
$|\sigma|\lesssim \Omega$ and we are considering $\beta\gtrsim
10$. Thus Eq. \ref{cond} is generally satisfied.
